### 2021-10-21 submission/response_to_reviewers.md

I am grateful that the reviewer sees value in this work and the questions I have raised. I apologize for taking so long to resubmit a revised version of the manuscript. In part because of the reviewers' comments, I paused work on this project to take on the question of whether the use of rarefaction was valid for alpha and beta diversity metrics. Two manuscripts: (1) a direct response to the original McMurdie and Holmes study and (2) an approach using a different simulation framework have now been posted as preprints to bioRxiv and submitted for review also at *mSphere* (see the new references **XX** and **XX**). Although the Editor's comments indicated that it was not necessary to take on the issue of rarefaction, I felt it would overcome many of the concerns that the reviewer and others have raised. Beyond the issue of rarefaction, I hope that the reviewer will find that I have addressed their comments throughout the manuscript and taken to revise the text to make my methods more clear and results more convincing. The line numbers included below correspond to the "track changes" version of the manuscript.

@@

**While these are important questions to address, I do not walk away from the manuscript - in its current form - convinced by its conclusions, especially given some of the strong language that is invoked. In particular, I think the manuscript could make a better case by improving its clarity in some areas, especially with respect to methodology and rationale, and by revising or adding analyses in other places to solidify conclusions. Moreover, some of the conclusions may require hedging given caveats that are not currently considered in the present analysis. Given that this manuscript upends long-standing practice and expectations, it seems especially important that it fill in these gaps.**

**Below, I itemize my criticisms with the manuscript. I hope Dr. Schloss sees these as constructive comments designed to help him improve this manuscript, which I ultimately think will play an important role in the advancement of microbiome data analytics.**

I appreciate the reviewers constructive comments and am confident that the manuscript is much improved because of their feedback. A significant challenge of this type of work is that we do not know the *true* answer for any of the original communities. One could use mock communities and risk over-fitting parameters to a highly artificial construct, which is my critique of the Bokulich et al. study (**XX**). Alternatively, one could try many methods on multiple studies and look for common results as was recently done by Nearing et al. (**XX**); however, the analyses cannot be easily controlled and the conclusions are unsatisfying. Finally, the approach I took was to design a thought experiment that asked what the data would look like if rare sequences were artifacts. I expected there to be more rare sequences when we had more sequences and there to be little overlap of rare sequences between different samples. As outlined in the section **The nature of singletons**, I found the opposite. The fraction of singleton ASVs decreased with increasing sequencing depth and there was considerable overlap of rare sequences in a poorly sequenced sample with the same sequences at greater abundance in more thoroughly sequenced samples.

There has been little to no rigorous work demonstrating that we *should* be removing low abundance or patchy sequences. The decision to employ abundance or prevalence based filtering has made by fiat rather than based on data. For example, McMurdie and Holmes decried removing samples and sequence data by rarefaction, but thought nothing of removing rare OTUs in their analysis (**XX**). Except for Bokulich et al. who justified abundance filtering because of low quality data, none of the papers describing methods for assigning sequences to ASVs experimentally justified their removal of rare sequences. Thus, I feel that I am left arguing against a "long-standing practice" with little objective and experimental support.

@@

**1. It is currently challenging to evaluate the work because the methodological details used to conduct it are frequently sparsely described. The materials and methods are in effect integrated into the results section. While I do not have a problem with this per se, it seems to have come at the cost of not providing specific details about various analytical steps that were implemented as part of the analysis as well as the rationale for implementing these techniques. This ranges from relatively straightforward aspects of methods, such as the reason for why these specific data sets were selected (from among the others that fit the same criterion of targeting the V4 locus with the same sequencing technology), to whether multiple tests correction was applied to statistical tests. In other places, the methodology is confusing as described, especially regarding the various null model sampling approaches (for example, see my specific comment about line 200 below). While I appreciate that Dr. Schloss has made all code available to readers, I think it's important to ensure that an understanding of the approach by a reader does not depend upon their ability to read code. I therefore recommend that the technical approach be more explicitly described in the manuscript.**

I appreciate the reviewer's concern about the mixing of methods throughout the Results section. Clearly, I needed to do a better job of describing those connections and I appreciate the reviewer for pointing out places that could be improved. The revised manuscript moves the details of the simulations to the Methods section and replaces that text in the Results section with more rationale for the simulation design. I have also tried to improve the description of the statistical tests and thresholds I used with the simulations. All the comparisons were binary within each study and did not require use of corrections for multiple comparisons.

There are certainly more datasets that could fit my criteria, but I felt that 12 diverse datasets were sufficient. I would be happy to include additional datasets that the Editor or reviewer recommend. Ultimately, the actual datasets that are used is not critical as I am only using the datasets to simulate various community structures. I feel that the description of why datasets were selected is sufficiently clear in the "Datasets" subsection of the Results section and the "Data curation and analysis" subsection of the Materials and Methods section.

@@

**2. I'm not sure that the manuscript provides evidence that rare sequences should be retained in an investigation, but rather documents the sensitivity of commonly applied analytical approaches to the decisions made about whether and how rare sequences should be incorporated. The former statement requires demonstrating that the inclusion of rare variants improves analytical accuracy. While the present analysis suggests that including rare variants impacts alpha- and beta-diversity metrics as well as the sensitivity of tests that utilize these metrics (according to the null model approach), it is not evident if the filtered or unfiltered data better reflects the underlying ecology. As a brief aside, Figure 5 offers the clearest insight into the potential impact of filtering on analytical accuracy, but it is not clear from the methodology how treatment groups were established and thus it is hard to follow Dr. Schloss' interpretation here.**

Inclusion of rare variants definitely improves analytical accuracy. The **The nature of singletons** subsection of the Results section highlights why I think that the rare sequences are largely robust. However, even if that were not the case, I still think that their removal would be problematic. I showed that the inter-sample variation as measured by the coefficient of variation increases with removal of rare sequences (**LXX**) and this impacts the ability to detect differences between samples (**LXX**). Furthermore, when sample sizes are confounded with treatment group removal of rare sequences exacerbates the false detection rate (**LXX**).

As mentioned above, a consistent challenge in studies such as mine is our lack of knowledge of the true structure of microbial communities or their underlying ecology. Although Mock community data can tell us what is spurious and what is real, their community structure is highly artificial and unlike anything we see in nature. Analysis of the original studies also possible but would leave the interpretation of what better reflects the "underlying ecology" to the investigator or an interpretative mess. Therefore, we are left to use model communities where we know their true structure and perform simulations based on those models. Of course, "all models are wrong, but some are useful" (George Box) and we are left to ask whether the models I've created here illuminate concerns that we should have with removing rare sequences. The models presented here do not incorporate sequencing errors per se, rather rare ASVs that may be good or spurious. What we find (and the reviewer notes) in Figures 4 and 5 is that regardless of the quality of the sequences, removing rare sequences causes problems for statistical power (Figure 4) and false detection of differences (Figure 5B). My models argue that retaining rare ASVs provides a better sense of the ecology of the study because it makes it easier to detect real differences (i.e., power) and harder to detect false differences.

@@

**In general, much of the interpretation of the results appears predicated on the idea that the rare variants represent real taxa. However, there is only limited evidence in support of this view. In the section "The nature of singletons", Dr. Schloss finds that singletons are more prevalent in more shallowly sequenced samples and that singletons in a sample are often also present in another sample. Dr. Schloss concludes that these singletons are thus likely real variants and the rest of the investigation seems somewhat predicated upon this interpretation. However, I am left with several questions:**

**(A) Are the correlations meaningfully representative of a trend across all samples, or are they driven by a small number of outlier samples? Providing support of these trends by plotting the scatterplots for each study would help assuage this concern.**

I appreciate the reviewer's concern that the correlations presented in Figure 1AC could be driven by outliers. It is worth noting that I used Spearman correlation coefficients, which are robust to the effects of outliers since it is a ranked-based test. As suggested, by the reviewer, I created two new figures, Figures S2 and S3, which shows the requested plots for the data in Fig 1AC. There are no obvious outliers in these figures that are skewing the Spearman correlation values.

@@

**(B) Despite these observations, couldn't these rare taxa include artifacts nonetheless and, if so, do real rare variants and artifacts contribute uniform information in all subsequent analyses? An effort to directly measure the impact of spurious sequences on analytical accuracy would help offset this concern.**

Rare taxa could certainly be artifacts. At L264-267 I discuss how even under the best of circumstances PCR artifacts like chimeras and sequencing errors will persist. As part of this discussion I described two strategies for dealing with lingering artifacts: rarefaction (L268-272) and making comparisons between treatment based on relative differences in diversity rather than absolute quantities (L273-283). Of course, ideally researchers would generate the best possible sequence data where reads fully overlap with each other. Using rarefaction gives every sample an equal chance to have artifacts. As I emphasize in the Discussion, by keeping interpretation to relative differences and not inflating the importance of rare taxa we can further mitigate against problems caused by artifacts.

**(C) Since the processes underlying the generation of artifacts are not explicitly described, the reader is left unsure about the expectation that the formation of singleton artifacts is driven by sequencing depth.**

I hope that the revised text at **LXX** (2nd paragraph of Discussion) addresses the reviewer's concerns.

@@

**3. I'm not sure that the description of variant filtering accurately reflects common practice. For example, while many groups appear to apply count thresholds to determine which taxa should be retained for analysis, these thresholds are typically applied across all samples and are often cross referenced with prevalence filtering. As a result, taxa that are singletons in one sample but present in others are typically included in an analysis despite otherwise being below the count threshold. In Dr. Schloss' study, however, such taxa are considered candidates for filtration, which raises questions about the applicability of the results. Moreover, frequent practice thresholds on relative abundance rather than direct counts to offset the potential for sequencing depths to drive singleton formation effects. Dr. Schloss may consider discussing or even experimentally measuring the effect of these alternative strategies on the outcomes of his analyses.**

* As outlined in the Introduction, sample-by-sample abundance filtering is the norm for dada2, Deblur, and the Bokulich pipeline whereas study-wide abundance filtering is used by UNOISE2.
 Closed reference clustering as a filtration method...

*   [ simulation? ]
  - study-wide abundance filtering 
  - counting the number of samples that have a sample

* The reviewer suggests using relative abundance thresholds as an alternative to abundance thresholds. Relative abundance thresholds are effectively the same as rarefied OTU counts since, mathematically, relative abundances are the average number of counts for the OTU across a large number of resamplings divided by the total number of sequences in a sample. Unfortunately, filtering by relative abundances do not control for the total number of sequences in a dataset and so any ASV below the threshold will effectively have its abundances redistributed across the other ASVs. This is the same problem I encountered when removing rare ASVs. [ simulation? ]
* Fully overlapping read pairs have previously been recommended for 16S rRNA analyses, since the increase in sequencing errors in nonoverlapping regions can result in spurious OTU calls (11). Related to this suggestion, there has also been some concern about the base-quality performance of Illumina’s v3 kit chemistry toward the ends of the 300-nucleotide (nt) reads (20). These are valid concerns, and we have tweaked our quality-filtering steps to help minimize these problems. The distribution of quality scores in forward and reverse reads is shown in Fig. 3A and B to help visualize this issue. For both forward and reverse reads, there is a decrease in quality near the 3′ ends. After stitching these 300-nt reads together (Fig. 3C), there is a clear increase in quality in the overlapping region (from ~150 to 300 bp). Notably, there is a large degree of variation in quality, especially toward the end of the reads. After quality filtering, there is much less variation (Fig. 3D). Despite this improvement in quality score distributions, spurious OTU calls still occur, as shown by the massive number of raw OTUs originally called (139,253). After eliminating OTUs that are called by <0.1% of reads (our “remove low-confidence OTUs” step), which is the maximal expected bleed-through between MiSeq runs according to Illumina (21), we retained 4,504 OTUs. This 97% removal of (most probably) spurious OTUs demonstrates, along with the mock community results in Fig. 2, that reasonable OTU estimates can be achieved by using proper quality control. journals.asm.org/doi/full/10.1128/mSystems.00127-16
* UPARSE removes singletons?

@@

**4. Related to the above, it is not clear if the results presented here are generalizable to other investigations. For example, it is possible that the use of the high-quality data sets included in this analysis or the specific bioinformatic techniques applied here results in a lower than typical rate of artifact formation compared to what many readers may encounter in their own analyses. For example, studies that do not apply library prep approaches that produce two overlapping V4 reads, but that rather amplify a longer amplicon target and thus have lower sequence accuracy, may have an increased rate of forming artifacts in their data. Likewise, ASVs in this study are sequences that tolerate a 2 bp difference, which may differ from the set of ASVs produced using another approach. Since the study never directly measures the cost of including artifactual sequences (2B above), it is not clear if the recommendations made here are meaningful for investigations that involve data subject to a greater rate of artifact formation.**

I think the reviewer may be putting too much emphasis on the importance of the individual datasets. These datasets were selected to help me simulate communities with different properties, number sequences per sample, and number of samples per study. At the same time, I would resist any attempt to normalize the generation of data that we know has a high error rate. Why shouldn't researchers be expected to generate the best data they can?

@@

**5. Conclusions around rarefaction need more justification or support. There are several strongly worded statements in the manuscript about the use of rarefaction techniques, but there appears to be little direct assessment of the impact of rarefaction on study outcomes. Additionally, there is no comparison with alternative rarefaction-free normalization techniques, including simply filtering based on relative abundance. Given the on-going debate in the literature about the value of rarefaction, it seems important that the manuscript solidify this point if the language is to be included as is.**

As I mentioned in the preamble to my response to the reviewer's comments, since receiving their comments, I have written two manuscripts using different simulation approaches to demonstrate that rarefaction is critical for drawing proper inferences from microbiome studies. As outlined in the response to McMurdie and Holmes's study, their approach was effectively designed to make look rarefaction look bad - this was principally achieved through (1) removing 15% of the samples and then penalizing the method for the loss of those samples and (2) basing the analysis on a cluster analysis of samples that used a clustering algorithm that performed the worst of the available options. In the second manuscript, I developed a different simulation strategy that is similar to the one used in this study and employed additional normalization approaches. Both studies show that rarefaction is critical for controlling for uneven sampling effort when pursuing alpha and beta diversity analyses. I contend that the results I have shown regarding coefficients of variation and type 1 and 2 errors support the use of rarefaction. Finally, the reviewer encourages the use of relative abundance, which I show in my rarefaction papers can lead to problematic interpretations.

@@

**6. The rationale underlying some of the analyses, especially with respect to the null models, needs support to help readers connect the results with the conclusions. For example, for the section on line 122 ("The impact of removing rare ASVs on the information represented in each sample.") it seems self-evident that removing rare taxa will affect the estimate of alpha-diversity, especially when considering the well documented rank abundance distribution and long-tail of low abundance taxa that typically comprises microbial communities. Dr. Schloss may get more mileage out of this section by clarifying why this analysis is noteworthy, or by focusing of metrics that are less directly sensitive to the number of taxa (perhaps evenness).**

* Sobs is definitely obvious
* Bigger point in Figure 3 is that the coefficient of variation between samples goes up with removal of rare sequences reducing power to detect differences in communities
* Will also calculate Shannon evenness (keep in mind that this is dependent on richness)

@@

**Additionally, I'm not sure that the treatment effect simulation analysis is as meaningful as indicated by the text (though again, I'm not sure I follow the approach properly). If the ASV sampling procedure to determine which ASVs are increased in abundance is simply selecting from a list of ASVs, then the long-tail of rare ASVs will tend to be selected versus those that tend to abundant. Consequently, when the procedure amplifies the abundance of these (and only these) selected ASVs, again which are biased towards rare taxa, then it will be rare taxa driving the differences between simulated treatments. Abundance-based filtering of said taxa will obviously have an effect on the analysis. Is this akin to what happens in microbial communities in nature? Well, maybe in some instances, but lots of studies find that it's the abundant taxa that differentiate treatments. Such studies may not need the singleton ASVs to observe the effect. Could those ASVs help clarify the effect? Perhaps. But inclusion of artifact sequences (which again are not directly tested here) could theoretically increase the error of the analysis in such cases at a greater rate. I would recommend considering a parallel analysis that simulates ASVs based on their abundance to help clarify whether the interpretation drawn from this section is relevant to communities that behave along the lines described above. Alternatively, hedge the conclusion by clarifying the above caveat, but that this analysis may yet reveal insights into how rare ASVs could reveal differences currently being missed in communities that do not substantially change with respect to the abundant fraction of community membership.**

* As the reviewer points out, we don't know. There will certainly be times where differences between communities are driven by rare populations and times where it splits by abundant populations. We really don't know. Those that are driven by abundant populations will be obviously different already so it's a moot point for that case
* High quality data reduces risk of artifacts and rarefaction allows us to control for inclusion of artifacts
* Not sure if reviweer is asking about differential abundance, which isn't my concern here

@@

**Minor comments:**

**Line 94: What is meant by "relatively small" when referring to the marine and soil data sets? Based on table 1, it seems to be the number of samples. But you can help readers by being more explicit here especially given that the surrounding context of discussion pertains to sequencing depth.**

The reviewer is correct that I was referring to the number of samples. I have clarified this sentence.

@@

**Line 102: "In addition, with the exception of the samples from the marine and sediment datasets (Spearman correlation, P>0.05)..." Indicate parenthetically here that these samples were non-significant (as opposed to the immediate assumption that they are positively associated) and move the parenthetical pointer to spearman and the pvalue to later in the sentence.**

The sentence has been edited as suggested by the reviewer.

**Line 148: "For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples." It would be more compelling to demonstrate that this expectation was met in the resulting distributions. Can you offer graphical or statistical evidence in support of this?**

This was shown in Figure 3 when rare sequences were not removed (i.e. when the smallest number of sequences per ASV was equal to 1). In this case the coefficients of variation were nearly zero. Furthermore, in Figure 5A and 5B the Type 1 errors were near the expected 5% (i.e., P=0.05) for the case when the smallest number of sequences was equal to 1.

@@

**Line 175: How many samples are being selected for each simulated treatment in this analysis?  At first I thought all samples in a study were selected, but it looks like this isn't the case, as an undescribed number of "communities were randomly sampled from the null distribution..."**

The simulations did indeed use all of the samples. I have clarified the text to indicate that all of the samples were randomly assigned into one of two treatment groups

@@

**Line 179: What p-value threshold is being applied to "detect a difference" and quantify the "fraction of significant tests"?**

I have clarified this sentence to indicate that a significant test was one where the P-value threshold was 0.05.

### 2021-10-21 submission/manuscript.Rmd
Address reviewer's comments

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE}
options(tidyverse.quiet = TRUE)

@@

Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform. In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013]. In addition, a polishing step is often used to identify ASVs based on the frequency and similarity of sequences [@Kozich2013; @Edgar2016; @Amir2017]. In a second approach, the quality scores and types of errors are modelled to cluster sequence reads directly into ASVs [@Callahan2016]. Regardless of the approach, many pipelines advocate for abundance-based screening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016]. Some algorithms recommend removing all ASVs that appear one (i.e. singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016]. A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013]. After assigning sequences to ASVs, ASVs are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.

The abundance-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs. Sequencing of mock communities confirms that artifacts tend to be rare [@Kozich2013; @Bokulich2013]. Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed [@Bokulich2013]. However, this approach effectively overfits the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples. It is necessary to think more deeply about the practice of abundance-based screening.

The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.  These recommendations appear to assume that the number of reads per sample is consistent within and between studies. However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g. Table 1 and Figure S1). An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. However, rarity, both in terms of prevalence and incidence, is an important ecological concept [@Reid2011]. Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

In the current study, I used published sequence data from `r nrow(table_1)` studies to investigate the nature of rare ASVs (i.e. those that appear 10 or fewer times) and the effect that removing them has on downstream analysis of microbial communities. The analysis was also performed using traditional OTUs, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other. The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1; Figure S1). After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples. The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)`. Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.

@@

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. If such ASVs are artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, the median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). In addition, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets (Spearman correlation, P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2). This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact. This suggests that the singleton ASVs are not as likely to be artifacts as previously thought. Furthermore, if singleton ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B). Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively ((Figures 1C and S3)). The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes more non-artifact ASVs than those that are artifacts.

@@

**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed. This effectively made every community in a study a statistical sample of the study-wide composite community distribution. For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples. Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study. The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S5). The null distribution data were used in the remainder of the study to minimize the risk of bias.

@@

b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.))
b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n())
b_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% slice(1, n())

@@

b_base_beta <- b_beta %>% filter(clustering == "pc") %>% arrange(`1`)

@@

\includegraphics{figure_1.png}

@@

\includegraphics{figure_2.png}

\includegraphics{figure_3.png}

\includegraphics{figure_4.png}

>>>>

```{r knitr_settings, tidy=TRUE, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE}
options(tidyverse.quiet = TRUE)

@@

Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform. In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013]. In addition, a polishing step is often used to identify ASVs based on the frequency and similarity of sequences [@Kozich2013; @Edgar2016; @Amir2017]. In a second approach, the quality scores and types of errors are modelled to cluster sequence reads directly into ASVs [@Callahan2016]. Regardless of the approach, many pipelines advocate for abundance-based screening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016]. Popular algorithms employ abundance-based filtering of sequences that appear one [i.e. singletons, DADA2, @Callahan2016], four [UNOISE2, @Edgar2016], or ten [Deblur, @Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of counts from across the samples [@Edgar2016]. Beyond these abundance-based approaches, researchers will also often employ a prevalence-based filter. For example, Nearing et al. [@Nearing2022] used Deblur to produce ASVs [@Amir2017], which removes any sequences that appear 10 or fewer times in each sample and required that an ASV appear in 10% of samples seemingly regardless of their abundance in those samples. A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013]. Once sequences are assigned to ASVs, ASVs are often analyzed as their own taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.

The abundance and prevalence-based screening approach assumes that rare ASVs are more likely to be artifacts than more abundant ASVs. Sequencing of mock communities confirms that artifacts tend to be rare [@Kozich2013; @Bokulich2013]. Proponents of abundance-based screening point to their ability to obtain the correct number of ASVs, OTUs, or phylotypes with data generated from sequencing mock communities when rare ASVs are removed [@Bokulich2013]. However, this approach risks overfitting the curation pipeline to data generated from a phylogenetically simple community with an atypical community distribution that is often sequenced to a coverage that is not achieved with biological samples. It is necessary to think more deeply about the practice of abundance-based screening and question whether all rare or patchily-distributed sequences are artifacts. Rarity and patchiness are important ecological concept that appear to make microbial ecologists uncomfortable.

The minimum abundance thresholds that have been proscribed were largely developed and applied without regard for the number of sequences generated from each sample. These recommendations appear to assume that the number of reads per sample is consistent within and between studies. However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g. Table 1 and Figure S1). An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. One exception to the use of threshold counts was proposed by Bokulich et al. [@Bokulich2013] who suggested a 0.005% relative abundance threshold; however, they cautioned that this threshold was dependent on the quality of the data and should be reassessed with mock community data. Rather than removing rare or patchy ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled and problems inherent in using relative abundance, scaling, and other normalization procedures are avoided [@Schloss2023rare; @Schloss2023wnwn].

Experience sequencing biological samples also demonstrates that there are *bona fide* ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. For example, my research group frequently encounters cases where it is possible to culture *Clostridiodes difficile* from human and mouse fecal samples, but not find their 16S rRNA gene sequences or to find their sequences early in colonization but not once the organism has been cleared [e.g. @Lesniak2022; @Schubert2014]; naturally, we only find their sequences when the plating density is the highest. This seeming patchiness is a product of their ecology and not an artifact. Rarity, both in terms of abundance and prevalence, is an important ecological concept [@Reid2011]. Removing rare ASVs likely hinders one's to ability to make biological inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

In the current study, I used sequence data from `r nrow(table_1)` published studies to investigate the nature of rare ASVs (i.e. those that appear 10 or fewer times) and the effect of removing them on downstream analysis of microbial communities. The analysis was also performed using traditional OTUs, where ASVs subjected to abundance-based screening were clustered such that the ASVs within an OTU were no more than 3% different from each other. The results reject the assumptions built into abundance-based screening and highlight the problems inherent in removing rare ASVs.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1; Figure S1). After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples. The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)`. Strikingly, aside from the `r oxford_comma(low_fold_samples)` datasets, which had a relatively small number of samples, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.

@@

**The nature of singletons.** Removal of rare ASVs is commonly justified as a method of removing ASVs that are artifacts. The median percentage of sequences that were discarded when singleton ASVs were removed from each dataset varied between `r oxford_comma(percent_singletons_range_vals)`% (`r oxford_comma(percent_singletons_range_names)`). If these singleton ASVs were artifacts, then one would expect the number of singleton ASVs to accumulate with sequencing depth. Contrary to this expectation, with the exception of the samples from the `r oxford_comma(cor_n_nosig)` datasets whose Spearman correlation was not significantly different from 0.00 (P>0.05), the fraction of singleton ASVs in samples was negatively correlated with the number of sequences in each sample with a range between `r oxford_comma(cor_n_sig_values)` (`r oxford_comma(cor_n_sig_names)`) (Figures 1A and S2). This showed that with additional sequencing, the probability of seeing singleton ASVs in multiple samples was greater than the probability of generating an artifact. This suggests that the singleton ASVs are not as likely to be artifacts as previously thought. Furthermore, if singleton ASVs were artifacts, then one would not expect to find them in other samples from the same dataset. In fact, singleton ASVs from samples with fewer sequences were often found in samples with more sequences. At least 50% of the singleton ASVs found in the samples from the `r oxford_comma(high_coverage)` datasets were found in another sample from the same dataset (Figure 1B). Considering the likelihood of finding an ASV duplicated in another sample is confounded by the number of samples and inter-sample diversity, the high coverage of singleton ASVs in these datasets was remarkable. The correlation between the number of sequences in a sample and the fraction of that sample's singleton ASVs that were covered by another sample in the dataset was significant and negative for `r nrow(coverage_n) - length(coverage_n_nosig)` of the datasets ranging between `r oxford_comma(coverage_n_sig_values)` for the `r oxford_comma(coverage_n_sig_names)` datasets, respectively ((Figures 1C and S3)). The negative correlation indicated that the singleton ASVs in the smaller samples were more likely to be covered by ASVs in the larger samples. Among the three datasets without a significant correlation (Spearman correlation, P>0.05), the marine and soil datasets had the fewest samples in our collection and the stream dataset already had a high level of coverage regardless of the number of sequences. Contrary to the common motivation for removing rare ASVs, these results indicate that this practice disproportionately impacts samples with fewer sequences and likely removes more non-artifact ASVs than those that are artifacts.

@@

**Removing treatment group effects from community data.** One challenge with using the observed ASV and OTU frequency data to investigate the effects of removing rare ASVs is that we do not know the true difference between treatment groups to know whether the false detection rate or statistical power have been altered by their removal. Therefore, I used the observed frequency data to simulate samples where I could specify the true difference between the treatment groups. To minimize the effects of the random number generator seed, I replicated the analysis 100 times. The first simulation I created was a null model where all of the samples in a dataset were drawn from the same sampling distribution, without replacement. When I repeated the analysis investigating the effects of removing rare ASVs on richness, diversity, and information loss the losses were larger than were detected with the observed data (Figure S5). This result was likely due to the reduced inter-sample variation and patchiness in the datasets.

@@

b_base_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.)) %>%
	rename('one' = `1`)

b_base_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% arrange(`1`) %>% slice(1, nrow(.)) %>%
	rename('one' = `1`)

b_pc_sobs <- b_alpha %>% filter(metric == "sobs", method == "pc") %>% slice(1, n())
b_pc_shannon <- b_alpha %>% filter(metric == "shannon", method == "pc") %>% slice(1, n())

@@

b_base_beta <- b_beta %>% filter(clustering == "pc") %>% arrange(`1`) %>%
	rename('one' = `1`)

@@

@@
